Why do we Eval?\\[2mm]\Large A large-scale study of Eval usage in R

### 2020-11-27 inst/paper/main.tex
sec 3

Wang et al.~\cite{wang} analyzed use of dynamic features in 18 Python
programs to find if they affect file change-proneness.  Files with
dynamic features are significantly more likely to be the subject of
changes, than other files.  Chen et al. looked at the correlation
between code changes and dynamic features, including \eval, in 17
Python programs~\cite{chen}. They did not observe many uses of \eval.
Callau et al.~\cite{oscar} performed an empirical study of the usage
of dynamic features in 1,000 Smalltalk projects. While \eval itself is
not present, Smalltalk has a rich reflective interface that gives
programmer great flexibility. Their work was based on a static
analysis.  %% TODO:: Say someting meaningfull about OSCAR

@@

\begin{figure}[!t]\hspace{-5mm}\includegraphics[width=1.05\linewidth]
{pipeline.pdf}\caption{Analysis Pipeline}\label{fig:pipeline}
\end{figure}

\section{Methodology}

@@

Our infrastructure consists of a dynamic analysis pipeline to record the
calls to \eval and RMarkdown notebooks for data analysis and visualization.
Figure~\ref{fig:pipeline} shows the main steps of the pipeline. We provide
the runtime, input and output data size for each step. All runs are done on
a single server Intel Xeon 6140, 2.30GHz with 72 cores and 256GB of RAM.

The pipeline begins with downloading and installing open source R packages
from CRAN for the corpus along with their dependencies. The Kaggle programs
are downloaded from \href{http://www.kaggle.com}{Kaggle} separately by a
crawler.  Next, code size and coverage metrics from the installed packages
are extracted using the \href{ https://github.com/r-lib/covr}{\covr}
package. This is followed by extraction of runnable programs from R
packages: tests, examples and vignettes as well as from Kaggle. The code of
each extracted program is wrapped into a call to dynamic analyzer
\emph{evil} (\emph{Ev}al \emph{I}nspection \emph{L}ibrary) that collects eval
usage. The reason for this is that we only want to record eval usage for the
extracted code. Without this, the data would include eval calls from the
unit testing frameworks as well as from bootstrapping R virtual machine
itself. To avoid any interference, each program is run in its own R
instance. The \emph{evil} framework is implemented as a R package in 2K lines of
R and 400 lines of C++ code.

The dynamic analyzer builds upon the dynamic analysis framework, \instrumentr
that we have implemented to enable us to write dynamic analysis logic in R. It
is an R package implemented in 2.5K lines of R and 6K lines of C++ code. It
internally uses a modified R interpreter, \rdyntrace~\cite{oopsla19a}, that
exposes hooks from within the interpreter implementation for events of interest.
\instrumentr serves as an intermediary between \rdyntrace and \evil, it
intercepts the hooks exposed by \rdyntrace and attaches R functions exported by
\evil as callbacks. The \evil callbacks execute on corresponding interpreter
events.

All steps of this pipeline are parallelized using GNU
parallel~\cite{GNUparallel} and orchestrated by GNU make. To schedule and
parallelize extraction and analysis of programs, we use the \runr
package. Furthermore, \runr gracefully handles and reports failures across
large-scale program runs which greatly aids debugging of the analysis
pipeline.  The data extracted by \evil from each program is concatenated,
cleaned and summarized in the post-processing phase by custom R
scripts. Finally, the summarized data is analyzed in RMarkdown notebooks to
gather insights. Apart from the figures, the data points included in the
paper are also generated by RMarkdown notebooks as latex macros.

In this study, we report on a corpus of \CorpusCorePackages R core and
\CorpusPackages CRAN packages with \CorpusAllCodeRnd lines of source code
(native + R).  We run \CorpusAllProgramsRnd programs extracted from those
packages as well as from the user-written data-analyses shared on the Kaggle
platform. In this section we describe this corpus and methodology used to
assemble it.

\subsubsection{CRAN Packages}

We have selected the TOP \CorpusPackages packages based on their reverse
dependencies, \ie how many other packages depend on them.\footnote{Extracted
  from package metadata using a builtin function.}  This helps us find
packages which should be better written and have better code coverage.  The
resulting set contains \CorpusRCodeRnd lines of R code and
\CorpusNativeCodeRnd lines of native code.\footnote{Excluding comments and
  blank lines using \url{github.com/AlDanial/cloc}.}  For each package we
use its runnable code to compute the package code coverage.\footnote{Computed
  using \url{ https://github.com/r-lib/covr}.} On average, across the
\CorpusPackages packages, it is \CorpusMeanExprCoverage.
Figure~\ref{fig:corpus} shows these packages, the size of the dots reflects
the project's size in lines of code.  The x-axis indicates code coverage in
percents and the y-axis gives the number of call sites to \eval that were traced, in log
scale. Dotted lines indicate means. Packages with over
\CorpusEvalsPackageTreshold eval call sites are named.

All included packages come from the Comprehensive R Archive Network
(CRAN\footnote{\url{http://cran.r-project.org}}), the largest repository of R
code with over \CorpusAllCranRnd packages.\footnote{CRAN receives about 6 new
  package submissions a day~\cite{Ligges2017}.} Unlike other open code
repositories such as GitHub, CRAN is a curated repository. Each submitted
package must abide to a number of well-formedness rules that are automatically
checked asserting certain quality. Most relevant for this work is that all of
the \emph{runnable code} is tested and only a successfully running package is
admitted in the archive. This is important as we use this runnable code for the
dynamic analysis of eval calls.

There are three sources of runnable code in a R package. Next to the package
\emph{tests},\footnote{We had to exclude most tests because the {testthat}
  harness uses \eval and thus causes the entire test to register as an \eval
  call.} there are also \emph{examples} and \emph{vignettes}. Examples are R
code snippets extracted from package documentation into scripts files.
Vignettes are long-form description of package functionality written using
literate-programming style in \LaTeX\xspace or Rmarkdown with chunks of R
code that can be similarly extracted into R scripts. R provides builtin
functions for tangling the embedded code into files.  The selected packages
contain \CorpusPackagePrograms programs with \CorpusPackageProgramsCodeRnd
lines of R code: \CorpusExamplesProgramsRnd examples with
\CorpusExamplesCodeRnd lines of code, and \CorpusVignettesProgramsRnd
vignettes.

\subsubsection{Kaggle Scripts}

To represent end-user code in the corpus, we turned to Kaggle, an online
platform for data-science and machine-learning. The website allows people to
share data-science competitions, a data-analysis problem together with data,
for which users try to find the best solution. The solutions, called
{kernels}, are then shared back to the platform in the form of either plain
scripts or as notebooks.  One of the most popular competition is about
predicting passenger survival on
Titanic\footnote{\url{https://www.kaggle.com/c/titanic}} with \CorpusKaggle
kernels in R (over 1/4 of all available R kernels) which we used for our
corpus.

Unlike CRAN, Kaggle is not a curated repository and therefore there are no
guarantees about the quality of the code. After downloading all of the
\CorpusKaggle kernels and extracting the R code from the various
formats,\footnote{We use {\sf rmarkdown} to convert from notebooks to R.} we
found that \CorpusDuplicatedKaggle were whole-file duplicates. From
\CorpusRunnableKaggle kernels, \CorpusFailedKaggle failed to execute. Next
to various runtime exceptions, common problems were parse errors and
misspelled package names.  The final set contains \CorpusFinishedKaggle
kernels with \CorpusFinishedKaggleCodeRnd lines of R code.

\subsection{Threats to Validity}
The primary threat to validity is the issue of
code coverage, a cause of concern for any dynamic approach. We mitigate this
by including focusing on R packages that have high code coverage. We turn
off the bytecode compiler for this study. The bytecode compiler can also
call \eval. We do not get source locations for \UndefinedEvalsRnd eval
calls. In these cases \eval is either passed as an argument to a
higher-order functions or is defined in a function returned by a
higher-order function and the R parser does not retain location information
for \eval. However, this is a meager \PercentUndefinedEval of all eval calls
and is unlikely to affect our analysis. We ignore calls to the native \eval
function exposed by R. We also ignore the \c{rlang::tidy_eval} function
which uses native \eval internally because \c{rlang} is used to implement a
DSL for data analysis in R. It introduces a new first-class promise object
called \c{quosure} for which it implements special evaluation support in
\c{tidy_eval}.

>>>>

Wang et al.~\cite{wang} analyzed use of dynamic features in 18 Python
programs to find if they affect file change-proneness. Files with
dynamic features are significantly more likely to be the subject of
changes, than other files. Chen et al. looked at the correlation
between code changes and dynamic features, including \eval, in 17
Python programs~\cite{chen}. They did not observe many uses of \eval.
Callau et al.~\cite{oscar} performed an empirical study of the usage
of dynamic features in 1,000 Smalltalk projects. While \eval itself is
not present, Smalltalk has a rich reflective interface. The authors
found that reflective are used in less than 2\% of methods. The most
common reflective method is \c{perform:}; it send a message that is
specified by a string. These features are mostly used in the core
libraries.

Bodden et al.~\cite{bodden} looked at usage of reflection in the Java
DaCapo benchmark suite. They found that dynamic loading was triggered
by the benchmark harness. The harness then executes methods via
reflection, this caused static analysis tools to generate an incorrect
call graph for the programs in DaCapo.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Methodology}

This section explains our methodology for selecting the corpus that
will be analyzed and how we implemented our dynamic analysis.

Our corpus is assembled out of two set of programs, one consisting of
popular libraries obtained from the Comprehensive R Archive
Network\footnote{CRAN is the largest repository of R code with over
\CorpusAllCranRnd packages. It receives about 6 new package
submissions a day~\cite{Ligges2017}. Unlike other open code
repositories such as GitHub, CRAN is a curated repository. Each
submitted package must abide to a number of well-formedness rules that
are automatically checked asserting certain quality. Most relevant for
this work is that all of the runnable code is tested and only a
successfully running package is admitted in the archive.} (we refer to
it as {\bf CRAN}) and the other from the Kaggle data science
competition (written {\bf Kaggle}). The intent here is to contrast
code written by experienced developpers (CRAN) with code authored by
typical end users of the language (Kaggle).

All the code in our is runnable, i.e. we have scripts that invoke the
CRAN packages and the Kaggle programs have input data. Clearly code
coverage will vary; as is the case for any dynamic analysis, the
quality of results is predicated on how representative the executions
we can observe are with respect to all possible runs of the corpus. To
mitigate the threat to validity attached to limited coverage, we
obtain clients that will generate multiple \emph{runs} of CRAN
packages. In this context, a run is the execution of a CRAN package or
Kaggle program with a particular set of input values.

\medskip\noindent\emph{CRAN Packages.} The packages that are included
in this study are the top \CorpusPackages packages based on their
reverse dependencies. The set of reverse dependencies for some package
\emph{P} is transitive closure of packages that import \emph{P}. The
hypothesis underlying that choice is that packages with a large set of
reverse depencies are likely to be higher in quality and their tests
will provide better coverage. The packages are implemented in R with
some C and Fortran code. Using \c{cloc}, we measure a total
\CorpusRCodeRnd lines of R code and \CorpusNativeCodeRnd lines of
native code. For each package, we use its runnable code (tests, etc.)
to compute code coverage. On average, the cde coverage is
\CorpusMeanExprCoverage which is acceptable without being exhaustive.
Figure~\ref{fig:corpus} shows these packages, the size of the dots
reflects the project's size in lines of code. The x-axis indicates
code coverage in percents and the y-axis gives the number of call
sites to \eval that were traced, in log scale. Dotted lines indicate
means. Packages with over \CorpusEvalsPackageTreshold eval call sites
are named.

To analyze a package, we need client code that invokes its methods.
There are three sources of built-in runnable code that come with each
CRAN package: \emph{tests, examples} and \emph{vignettes}. They are,
respectively, traditional unit tests, code snippets from the
documentation, and long-form use-cases written in Rmarkdown. Examples
and vignettes are automatically extracted and turned into scripts,
their input is bundled with the package. Most tests had to be discared
due to a limitation of our pipeline: the \c{testthat} harness uses
\eval and thus causes the entire test to register as an \eval call.
The selected packages are bundled with \CorpusPackagePrograms
programs; \CorpusExamplesProgramsRnd examples and
\CorpusVignettesProgramsRnd vignettes.

\medskip\noindent\emph{Kaggle Scripts.} Kaggle is an online platform
for data-science and machine-learning. The website allows people to
submit data-analysis problems, users compete to find the best
solution. The solutions are uploaded to the platform as either plain
scripts or notebooks. We chose one of the most popular competition,
predicting the survival of passengers the
Titanic\footnote{\url{https://www.kaggle.com/c/titanic}}. Unlike CRAN,
Kaggle is not curated. After downloading the \CorpusKaggle solutions
and extracting the R code, we found that \CorpusDuplicatedKaggle were
duplicates. From the remaining \CorpusRunnableKaggle solutions,
\CorpusFailedKaggle failed to execute. Next to various runtime
exceptions, common problems were parse errors and misspelled package
names. The final set contains \CorpusFinishedKaggle programs
implemented in \CorpusFinishedKaggleCodeRnd lines of R code.

@@

The results presented in this paper are the result of an automated
analysis pipeline that acquires the code of packages, extract
metadata, executes programs, traces their behavior and summarizes the
observations. Figure~\ref{fig:pipeline} shows the main steps of the
pipeline along with approximate time to execute each step, the data
size, and the number of elements manipulated by the stage. Timings are
for runs on an Intel Xeon 6140, 2.30GHz with 72 cores and 256GB of
RAM.

\begin{figure}[!h]\hspace{-5mm}\includegraphics[width=1.05\linewidth]
{pipeline.pdf}\caption{Analysis Pipeline}\label{fig:pipeline}
\end{figure}

The first step in the pipeline consist of downloading CRAN packages
along with their dependencies and acquiring Kaggle programs with the
help of a web crawler. The second step, is to compute code size and
coverage metrics for the CRAN packages (we use the \covr package for
coverage). The third step consist in extracting runnable programs from
packages: i.e. the tests, examples and vignettes. Each extracted
program is wrapped into a call to our dynamic analyzer --- the tool is
called \emph{evil} for \underline{ev}al \underline{i}nspection
\underline{l}ibrary. This step is needed to ensure that we record
\eval usage only for the target package. Without this, the data would
include eval calls from the unit testing frameworks as well as from
bootstrapping R virtual machine itself. To avoid any interference,
each program is run in its own R instance.
The fourth step in the pipeline is to perform dynamic analysis for
each run of a CRAN package or Kaggle program.

The dynamic analyzer builds upon the dynamic analysis framework,
\instrumentr that we have implemented to enable us to write dynamic
analysis logic in R. \instrumentr serves as an intermediary between
\rdyntrace and \evil, it intercepts the hooks exposed by \rdyntrace
and attaches R functions exported by \evil as callbacks. The \evil
callbacks execute on corresponding interpreter events.

The data extracted by \evil from each program is
concatenated, cleaned and summarized in the post-processing phase by
custom R scripts. Finally, the summarized data is analyzed in
RMarkdown notebooks to gather insights. Apart from the figures, the
data points included in the paper are also generated by RMarkdown
notebooks as latex macros.

The \emph{evil} framework is implemented as a R package in 2K lines of
R and 400 lines of C++ code. \instrumentr is an R package implemented
in 2.5K lines of R and 6K lines of C++ code. It internally uses a
modified R interpreter, \rdyntrace~\cite{oopsla19a}, that exposes
hooks from within the interpreter implementation for events of
interest.

All steps of this pipeline are parallelized using GNU
parallel~\cite{GNUparallel} and orchestrated by GNU make. To schedule
and parallelize extraction and analysis of programs, we use the \runr
package. Furthermore, \runr gracefully handles and reports failures
across large-scale program runs which greatly aids debugging of the
analysis pipeline.

\medskip\noindent\emph{Limitations.}
The primary threat to validity is the issue of code coverage, a cause
of concern for any dynamic approach. We mitigate this by including
focusing on R packages that have high code coverage. We turn off the
bytecode compiler for this study. The bytecode compiler can also call
\eval. We do not get source locations for \UndefinedEvalsRnd \eval
calls. In these cases \eval is either passed as an argument to a
higher-order functions or is defined in a function returned by a
higher-order function and the R parser does not retain location
information for \eval. However, this is a meager \PercentUndefinedEval
of all \eval calls and is unlikely to affect our analysis. We ignore
calls to the native \eval function exposed by R. We also ignore the
\c{rlang::tidy_eval} function which uses native \eval internally
because \c{rlang} is used to implement a DSL for data analysis in R.
It introduces a new first-class promise object called \c{quosure} for
which it implements special evaluation support in \c{tidy_eval}.
